---
title: "Mediations for count data"
output: html_document
date: "2023-11-16"
created by: "Thuy Pham"
---


Package {maczic} Mediation Analysis for Count and Zero-Inflated Count Data
```{r}
install.packages("maczic")
library(maczic)
```


Package bruceR: BRoadly Useful Convenient and Efficient R functions
```{r}
## Method 1: Install from CRAN
install.packages("bruceR", dep=TRUE)  # dependencies=TRUE
library(bruceR)

## Method 2: Install from GitHub
install.packages("devtools")
library(devtools)
devtools::install_github("psychbruce/bruceR", dep=TRUE, force=TRUE)
library(bruceR)
```


Option 1
```{r}
# Generate example data
n <- 50
# Generate IV as a text variable, 0-100%
x <- sample(0:100, n, replace = T)
# Generate moderator as a text variable, 0-100%
w <- sample(0:100, n, replace = T)
# Generate mediator as a continuous variable
m <- rnorm(n)
# Generate continuous covariate variable
cov <- rnorm(n)
# Generate count outcome
y <- sample(0:100, n, replace = T)

data <- data.frame(x, w, m, cov, y)

# Hayes' Model 8

## Fitted model for mediator
model.m <- glm(m ~ x*w, data = data) # here the mediator is a continuous variable and can be negative, so it is not a negative binomial or Poisson distribution here. If the mediator is also input as percentage, then model.m <- glm.nb(m ~ x*w, data = data) or model.m <- gam(m ~ x*w, data = data, family = nb()).

## Fitted model for outcome 
model.y <- glm.nb(y ~ x*w + m, data = data)

## Estimation via nonparametric bootstrap 
# install.packages("maczic")
# library(maczic)
?mediate_zi() # read more to understand all the parameters in this function.

fit <- mediate_zi(model.m, model.y, sims = 100, treat = "x", mediator = "m", covariates = "cov", 
                  boot = TRUE # A logical value. if 'FALSE' a quasi-Bayesian approximation is used for confidence intervals; if 'TRUE' nonparametric bootstrap will be used.
                  )
summary(fit)
bruceR::med_summary(fit)
```
Example output:

Causal Mediation Analysis 

Nonparametric Bootstrap Confidence Intervals with the Percentile Method

(Inference Conditional on the Covariate Values Specified in `covariates')

                         Estimate 95% CI Lower 95% CI Upper p-value
ACME (control)            0.00210     -0.07020         0.08    0.84
ACME (treated)            0.00197     -0.06952         0.08    0.84
ADE (control)            -0.14821     -0.67957         0.12    0.34
ADE (treated)            -0.14834     -0.67952         0.12    0.34
Total Effect             -0.14624     -0.72677         0.11    0.36
Prop. Mediated (control) -0.01438     -1.03148         1.05    0.92
Prop. Mediated (treated) -0.01349     -1.02771         1.05    0.92
ACME (average)            0.00204     -0.06986         0.08    0.84
ADE (average)            -0.14828     -0.67955         0.12    0.34
Prop. Mediated (average) -0.01394     -1.02959         1.05    0.92

Sample Size Used: 50 

Simulations: 100 


Mediation Analysis:
x ==> m ==> y

Conditional on ...

──────────────────────────────────────────────────
          Effect    S.E.   [Boot 95% CI]     p    
──────────────────────────────────────────────────
Indirect   0.002 (0.035) [-0.070, 0.081]  .840    
Direct    -0.148 (0.224) [-0.680, 0.116]  .340    
Total     -0.146 (0.223) [-0.727, 0.107]  .360    
──────────────────────────────────────────────────
Percentile Bootstrap Confidence Interval
Sample Size: 50
Simulations: 100 (Bootstrap)



#####################

Option 2: Using package mediation
```{r}
# Generate example data
n <- 50
# Generate IV as a text variable, 0-100%
x <- sample(0:100, n, replace = T)
# Generate moderator as a text variable, 0-100%
w <- sample(0:100, n, replace = T)
# Generate mediator as a continuous variable
m <- rnorm(n)
# Generate continuous covariate variable
cov <- rnorm(n)
# Generate count outcome
y <- sample(0:100, n, replace = T)

data <- data.frame(x, w, m, cov, y)

# Hayes' Model 8
## Fitted model for mediator
m1.m <- glm(m ~ x*w, data = data)

## Fitted model for outcome 
m1.y <- glm.nb(y ~ x*w + m, data = data)


med <- mediation::mediate(m1.m, m1.y,
                   treat = "x", mediator = "m",
                   sims=1000)
summary(med)
bruceR::med_summary(med)
```
Example of Output:


Causal Mediation Analysis 

Quasi-Bayesian Confidence Intervals

                         Estimate 95% CI Lower 95% CI Upper p-value
ACME (control)           -0.00966     -0.17080         0.12    0.87
ACME (treated)           -0.00968     -0.16949         0.12    0.87
ADE (control)            -0.20725     -1.07607         0.24    0.62
ADE (treated)            -0.20727     -1.07547         0.24    0.62
Total Effect             -0.21693     -1.09314         0.24    0.60
Prop. Mediated (control)  0.01668     -2.02255         1.80    0.88
Prop. Mediated (treated)  0.01737     -2.01656         1.79    0.88
ACME (average)           -0.00967     -0.17006         0.12    0.87
ADE (average)            -0.20726     -1.07577         0.24    0.62
Prop. Mediated (average)  0.01703     -2.01955         1.79    0.88

Sample Size Used: 50 

Simulations: 1000 


Mediation Analysis:
x ==> m ==> y
──────────────────────────────────────────────────
          Effect    S.E.   [MCMC 95% CI]     p    
──────────────────────────────────────────────────
Indirect  -0.010 (0.067) [-0.170, 0.116]  .870    
Direct    -0.207 (0.344) [-1.076, 0.243]  .616    
Total     -0.217 (0.360) [-1.093, 0.241]  .602    
──────────────────────────────────────────────────
Monte Carlo (Quasi-Bayesian) Confidence Interval
Sample Size: 50
Simulations: 1000 (Monte Carlo)





#####################

Option 3: Using fucntion PROCESS for mediation and/or moderation analyses within package BruceR

```{r}
# Generate example data
n <- 50
# Generate IV as a text variable, 0-100%
x <- sample(0:100, n, replace = T)
# Generate moderator as a text variable, 0-100%
w <- sample(0:100, n, replace = T)
# Generate mediator as a continuous variable
m <- rnorm(n)
# Generate continuous covariate variable
cov <- rnorm(n)
# Generate count outcome
y <- sample(0:100, n, replace = T)

data <- data.frame(x, w, m, cov, y)

# Hayes' Model 8
bruceR::PROCESS(data = data, y="y", x="x",
        meds= "m",
        mods="w",
        covs = "cov",
        mod.path= c("x-y", "x-m"),
        ci="mcmc", # Markov Chain Monte Carlo (Quasi-Bayesian)
        nsim=100, 
        seed=1)
```

Example output looks like this:


****************** PART 1. Regression Model Summary ******************

PROCESS Model Code : 8 (Hayes, 2018; www.guilford.com/p/hayes3)
PROCESS Model Type : Moderated Mediation
-    Outcome (Y) : y
-  Predictor (X) : x
-  Mediators (M) : m
- Moderators (W) : w
- Covariates (C) : cov
-   HLM Clusters : -

All numeric predictors have been grand-mean centered.
(For details, please see the help page of PROCESS.)

Formula of Mediator:
-    m ~ cov + x*w
Formula of Outcome:
-    y ~ cov + x*w + m

CAUTION:
  Fixed effect (coef.) of a predictor involved in an interaction
  denotes its "simple effect/slope" at the other predictor = 0.
  Only when all predictors in an interaction are mean-centered
  can the fixed effect denote the "main effect"!
  
Model Summary

────────────────────────────────────────────
             (1) y       (2) m    (3) y     
────────────────────────────────────────────
(Intercept)  52.940 ***   0.064   53.269 ***
             (4.704)     (0.139)  (4.707)   
cov          -1.882       0.170   -0.972    
             (5.253)     (0.156)  (5.369)   
x            -0.062       0.000   -0.022    
             (0.158)     (0.005)  (0.162)   
w                        -0.005   -0.040    
                         (0.005)  (0.166)   
x:w                       0.000    0.007    
                         (0.000)  (0.006)   
m                                 -6.456    
                                  (5.061)   
────────────────────────────────────────────
R^2           0.007       0.063    0.072    
Adj. R^2     -0.036      -0.020   -0.034    
Num. obs.    50          50       50        
────────────────────────────────────────────
Note. * p < .05, ** p < .01, *** p < .001.

************ PART 2. Mediation/Moderation Effect Estimate ************

Package Use : ‘mediation’ (v4.5.0), ‘interactions’ (v1.1.5)
Effect Type : Moderated Mediation (Model 8)
Sample Size : 50
Random Seed : set.seed(1)
Simulations : 100 (Monte Carlo)

Warning: nsim=1000 (or larger) is suggested!

Interaction Effect on "y" (Y)
─────────────────────────────
          F df1 df2     p    
─────────────────────────────
x * w  1.57   1  44  .217    
─────────────────────────────

Simple Slopes: "x" (X) ==> "y" (Y)
(Conditional Direct Effects [c'] of X on Y)
──────────────────────────────────────────────────────────────
 "w"           Effect    S.E.      t     p            [95% CI]
──────────────────────────────────────────────────────────────
 21.992 (- SD) -0.239 (0.211) -1.133  .263     [-0.665, 0.186]
 51.700 (Mean) -0.022 (0.162) -0.134  .894     [-0.348, 0.305]
 81.408 (+ SD)  0.196 (0.261)  0.750  .457     [-0.330, 0.722]
──────────────────────────────────────────────────────────────

Interaction Effect on "m" (M)
─────────────────────────────
          F df1 df2     p    
─────────────────────────────
x * w  0.12   1  45  .732    
─────────────────────────────

Simple Slopes: "x" (X) ==> "m" (M)
(Conditional Effects [a] of X on M)
──────────────────────────────────────────────────────────────
 "w"           Effect    S.E.      t     p            [95% CI]
──────────────────────────────────────────────────────────────
 21.992 (- SD) -0.001 (0.006) -0.240  .812     [-0.014, 0.011]
 51.700 (Mean)  0.000 (0.005)  0.056  .956     [-0.009, 0.010]
 81.408 (+ SD)  0.002 (0.008)  0.263  .793     [-0.013, 0.017]
──────────────────────────────────────────────────────────────

Running 100 * 3 simulations...
Indirect Path: "x" (X) ==> "m" (M) ==> "y" (Y)
(Conditional Indirect Effects [ab] of X through M on Y)
──────────────────────────────────────────────────────────────
 "w"           Effect    S.E.      z     p       [MCMC 95% CI]
──────────────────────────────────────────────────────────────
 21.992 (- SD)  0.014 (0.051)  0.279  .780     [-0.063, 0.146]
 51.700 (Mean) -0.003 (0.039) -0.086  .931     [-0.079, 0.071]
 81.408 (+ SD) -0.021 (0.065) -0.320  .749     [-0.186, 0.105]
──────────────────────────────────────────────────────────────
Monte Carlo (Quasi-Bayesian) Confidence Interval
(Effect, SE, and CI are estimated based on 100 Monte Carlo samples.)

Note. The results based on bootstrapping or other random processes
are unlikely identical to other statistical software (e.g., SPSS).
To make results reproducible, you need to set a seed (any number).
Please see the help page for details: help(PROCESS)
Ignore this note if you have already set a seed. :)


